Classless Java

### 2015-11-21 paper/sections/Formalization.tex
Section 2 (work in progress)
Commit in conflic, hope is easy to solve
Macros to make formalization auxiliary functions more coherents
Formalization more coherent

\begin{figure}[t]
\begin{grammar}
\production{
\e
}{
  \x\mid\MCall\e\m\es\mid\MCall{\C}\m\es\mid\MCall{\C\QM{.super}}\m\es\mid\x\QM=\e\QM;\e'\mid\obj
  }{expressions}\\
\production{
\obj
}{
\QM{new}\ \C\oR\cR\oC\fields\
\mh_1\oC\QM{return}\ \e_1\QM{;}\!\cC
\ldots
\mh_n\oC\QM{return}\ \e_n\QM{;}\!\cC
\cC
  }{object creation}\\
\production{\field}{\T\ \f \QM= \x\QM;}{field declaration}\\
\production{
\II
}{
 \ann\ \QM{interface}\ \C_0\ \QM{extends}\ \Cs\ \oC \methods\ \cC
  }{interface declaration}\\
\production{
\method
}{
 \QM{static}\ \mh\ \oC\QM{return}\ \e\QM{;}\!\cC
\mid
\QM{default}\ \mh\ \oC\QM{return}\ \e\QM{;}\cC
\mid
\mh\QM{;}
  }{method declaration}\\
\production{
\mh
}{
 \T_0\ \m\ \oR\T_1\ \x_1\ldots\T_n\ \x_n\cR
  }{method header}\\
\production{
\ann
}{
  \mixinAnn|\emptyset
  }{annotations}\\
\production{\Gamma}{
\x_1{:}\C_1\ldots\x_n{:}\C_n
}{environment}
\end{grammar}
\caption{Grammar of ClassLess Java}
\label{Grammar}
\end{figure}

@@

%% T-Invk
\inferrule[(T-Invk)]{
 \Gamma \vdash \e \in \C \\\\
\forall i\in 1..n\ \ \Gamma \vdash \e_i \in \_<:\C_i \\\\
  \textsf{mtype}(\m,\C) \!=\! \overline{\C} \!\!\to\! \C'
%\textsf{mmodifier}(\m,\C) \neq \textbf{static}
 }{
 \Gamma \vdash \e\QM.\m\QM(\overline{\e}\QM) \in \C' }
\quad\quad

%%T-StaticInvk
\inferrule[(T-StaticInvk)]{
\forall i\in 1..n\  \ \Gamma \vdash \e_i\in \_<:\C_i \\\\
\textsf{mtypeS}(\m,\C) \!=\! \overline{\C} \!\to\! \C'
%\textsf{mmodifier}(\m,\C) = \textbf{static}
}{
\Gamma \vdash \C\QM.\m\QM(\overline{\e}\QM) \in \C'}
\quad\quad

%%T-SuperInvk
\inferrule[(T-SuperInvk)]{
\Gamma(\this) <: \C \\\\
\forall i\in 1..n\ \ \Gamma \vdash \e_i\in \_<:\C_i \\\\
  \textsf{mtype}(\m,\C) \!=\! \overline{\C} \!\!\to\! \C'
%\textsf{mmodifier}(\m,\C_0) \neq \textbf{static} \\\\
}{\Gamma \vdash \C\QM.\QM{super}\QM.\m\QM(\overline{\e}\QM) \in \C'}

@@

%%T-Intf
 \inferrule[(T-Intf)]{
IT(\C) = \ann\ \QM{interface}\ \C\ \QM{extends}\ \overline{I} \
\oC\methods\ \cC\\\\
 \forall \QM{default}\ \mh\oC\QM{return}\ \e\QM;\cC \in \methods,
\ \ \Gamma^{\mh},\,\QM{this}{:}\C\vdash\e\in \_\subtype\C^{\mh} \\\\
 \forall \QM{static}\ \mh\oC\QM{return}\ \e\QM;\cC \in \methods,
\ \ \Gamma^{\mh}\vdash\e\in \_\subtype\C^{\mh} \\\\
\dom(\C)=\dom(\C_1)\cup\ldots\cup\dom(\C_n)\cup\dom(\methods)
 }{
\C \text{ OK}
}
\end{array}$
\caption{CJ Typing}
\label{ET}
\end{figure}

In Figure~\ref{ET} we show the typing rules.  We discuss first the
most interesting rules, that is \rn{t-Obj} and \rn{t-Intf}. Rule
\rn{t-Obj} is the most complex typing rule. Firstly, we need to
ensure that all field initializations are type correct, by looking up the type of
each variable assigned to a field in the typing environment and verifying that such type is a
subtype of the field type. Secondly, we check that all method bodies are
well-typed. To do this the enviroment used to check the method body
needs to be extended appropriately: we add all fields and their types;
add $\this:I$; and add the arguments (and types) of the respective
method. Finally, we need to check that all method headers are valid
with respect to the superinterfaces of $I$; and we also need to check
that all abstract and conflicted methods (that is methods, that need
to be explicitly overriden) in the superinterfaces have been
implemented. The definitions $\sigvalid$ and $\alldefined$ take care
of this: \\

$
\begin{array}{lll}
\sigvalid(\mh_1\ldots\mh_n,I) & = &
\forall i\in 1..n\ \ \mh_i\subtype\mBody(\m^{\mh_i},\C)\\

\alldefined(\mh_1\ldots\mh_n,I) & = &
\forall\m\mbox{ such that } \mBody(\m,\C)=\mh\QM; \exists i\in 1..n\ \m^{\mh_i}=\m
\end{array}
$\bruno{Marco, you probably want to double-check the explanation. Any
  more explanation needed?}
\bruno{Also, according to the text above, mbody returns the full
  method body, so in sigvalid, shouldn't we use mtype instead? The
  same for alldefined, should it be mtype? Still not very confident
  about alldefined. }

@@

The body of a method $\m$ contains all the relevant information with respect to that method, like the type of $\m$ as well as the modifier.
We use internally a special modifier $\conflicted$ to denote the case of two methods with conflicting implementation.\\*

$\mBody(\m,\C_0) = \override(\methods(\m),
\shadow(\m,\Aux{tops}(\Cs)))
$\\*
Where
$\metaVar{IT}(C_0) =
\ \ann\ \QM{interface}\ \C_0\ \QM{extends}\ \Cs \oC\methods\cC$

@@

\tops{} recover from the interface table only the ``needed'' methods, that is,
the non static ones that are not transitively reachable by following another, less specific, superinterface chain. Formally:\\*
$\method\in\tops(\m,\Cs) \mbox{ iff }\C\in\Cs\ ,\method=\mBody(\m,\C), \method\mbox{ not a static method}$\\*
${}_{}$\tab and $\nexists \C'\in\Cs\setminus\C$ such that $\C' \subtype \C$.

\shadow{} choose the most specific version of a method, that is the unique version available, or a conflicted version from a set of possibilities.
We do not model overloading, so it is an error if multiple versions are available with different parameter types. Formally:\\*
\!\!\!\!$\begin{array}{l}
%\shadow(\m,\C_1\ldots\C_n)=\shadow(\methods)
%\whereNote
% \mwhere\ \mBody(\m,\C_i)\in\methods
%\ \mif\ \mBody(\m,\C_i)\in\{\mh\QM;,\QM{default}\ \mh\mbox{\Q@\{return \_;\}@}\}
%\\ &\mor\ \mBody(\m,\C)=\mh\QM;
%\\

\shadow()=\none\\
\shadow(\method)=\method\\
\shadow(\overline{\mh\QM;})=\Aux{mostSpecific}(\overline{\mh\QM;})\\
\shadow(\methods)=\conflicted\mh\QM;
\whereNote
\mwhere\
\methods\mbox{ not of the form }\overline{\mh\QM;}\mbox{ and }
\Aux{mostSpecific}(\methods)\in\{\mh\QM;,\QM{default}\ \mh\mbox{\Q@\{return \_;\}@}\}\\
\Aux{mostSpecific}(\methods)=\method
\whereNote
\mwhere\ \method \in \methods\ \mand\ \forall \method' \in \methods :  \method \subtype
                                       \method', \\
\T\ \m\oR\T_1\x_1\ldots \T_n\x_n\cR \subtype \T' \m\oR\T_1\x_1'\ldots\T_n\x_n'\cR
\whereNote
 \mwhere\ \T\subtype \T'\\

\method \subtype
\QM{default}\ \mh\,\mbox{\Q@\{return \_;\}@}=
\method\subtype\mh\QM;
\\
\QM{default}\ \mh\,\mbox{\Q@\{return \_;\}@}\subtype\method=
\mh\QM;\subtype\method
\end{array}$
${}_{}$\\*
Where $\Aux{mostSpecific}$ return the most specific method using return type specialization as introduced in Java\marco{insert the version number}
We just check the subtype between method headers, so we discard information abut method implementation.\\*
${}_{}$\\*
The override function models how the implementation in an interface can override implementation in the superinterface; even in case of a conflict.
Note how we use the special value $\none$, and how (forth case) overriding can solve a conflict.
\\*
\!\!\!\!$\begin{array}{ll}
\override(\none,\none)=\none\\
\override(\method,\none)=\method\\

\override(\none,\method)=\method
\whereNote
\mwhere\ \method\mbox{ not of the form } \conflicted\ \mh\QM;\\
\override(\method,\method')
=
\method
\whereNote
\mwhere\ \method'\in\{\mh\QM;,\QM{default}\ \mh\,\mbox{\Q@\{return \_;\}@}, \conflicted\ \mh\QM; \},
\method\subtype\method'
\\
%\override(\method,\mh')
%=\method
%\whereNote
%\mif\ \method\in\{\mh\QM;,\QM{default}\ \mh\mbox{\Q@\{return \_;\}@} \}, \mh\subtype\mh'
%\override(\_,\_)=\error & Otherwise
%\textsf{shadow}(body_1, body_2)=\emptyset & \textsf{if }body_1.\textsf{modifier}=body_2.\textsf{modifier}=\textbf{static}\\
%\textsf{shadow}(body_1, body_2)=body_1 & \textsf{if }body_2=\emptyset\textsf{ or }body_2.\textsf{modifier}=\textbf{static}\\
%\textsf{shadow}(body_1, body_2)=body_2 \hspace{.1in}  & \textsf{if }body_1=\emptyset\textsf{ or }body_1.\textsf{modifier}=\textbf{static}
\end{array}$

@@

\marco{The notation mtype get only non static methods, the notation mtypeS get only static ones}

%\subsubsection{Interface and method Typing}

@@

%\lstinputlisting[linerange=5-10]{../UseMixinLombok/src/methodshadowing/test5/Test5.java} % APPLY:linerange=JAVABUG

>>>>

\begin{figure}[t]
\begin{grammar}
\production{
\e
}{
  \x\mid\MCall\e\m\es\mid\MCall{\C}\m\es\mid\MCall{\C\QM{.super}}\m\es\mid\x\QM=\e\QM;\e'\mid\obj
  }{expressions}\\
\production{
\obj
}{
\QM{new}\ \C\oR\cR\oC\fields\
\mh_1\oC\QM{return}\ \e_1\QM{;}\!\cC
\ldots
\mh_n\oC\QM{return}\ \e_n\QM{;}\!\cC
\cC
  }{object creation}\\
\production{\field}{\T\ \f \QM= \x\QM;}{field declaration}\\
\production{
\II
}{
 \ann\ \QM{interface}\ \C\ \QM{extends}\ \Cs\ \oC \methods\ \cC
  }{interface declaration}\\
\production{
\method
}{
 \QM{static}\ \mh\ \oC\QM{return}\ \e\QM{;}\!\cC
\mid
\QM{default}\ \mh\ \oC\QM{return}\ \e\QM{;}\cC
\mid
\mh\QM{;}
  }{method declaration}\\
\production{
\mh
}{
 \T_0\ \m\ \oR\T_1\ \x_1\ldots\T_n\ \x_n\cR
  }{method header}\\
\production{
\ann
}{
  \mixinAnn|\emptyset
  }{annotations}\\
\production{\Gamma}{
\x_1{:}\C_1\ldots\x_n{:}\C_n
}{environment}
\end{grammar}
\caption{Grammar of ClassLess Java}
\label{Grammar}
\end{figure}

@@

%% T-Invk
\inferrule[(T-Invk)]{
 \Gamma \vdash \e \in \C \\\\
\forall i\in 1..n\ \ \Gamma \vdash \e_i \in \_<:\C_i \\\\
  \textsf{mtype}(\m,\C) \!=\! \C_1\ldots\C_n \!\!\to\! \C'
%\textsf{mmodifier}(\m,\C) \neq \textbf{static}
 }{
 \Gamma \vdash \e\QM.\m\QM(\e_1\ldots\e_n\QM) \in \C' }
\quad\quad

%%T-StaticInvk
\inferrule[(T-StaticInvk)]{
\forall i\in 1..n\  \ \Gamma \vdash \e_i\in \_<:\C_i \\\\
\textsf{mtypeS}(\m,\C) \!=\! \C_1\ldots\C_n \!\to\! \C'
%\textsf{mmodifier}(\m,\C) = \textbf{static}
}{
\Gamma \vdash \C\QM.\m\QM(\e_1\ldots\e_n\QM) \in \C'}
\quad\quad

%%T-SuperInvk
\inferrule[(T-SuperInvk)]{
\Gamma(\this) <: \C \\\\
\forall i\in 1..n\ \ \Gamma \vdash \e_i\in \_<:\C_i \\\\
  \textsf{mtype}(\m,\C) \!=\! \C_1\ldots\C_n \!\!\to\! \C'
%\textsf{mmodifier}(\m,\C_0) \neq \textbf{static} \\\\
}{\Gamma \vdash \C\QM.\QM{super}\QM.\m\QM(\e_1\ldots\e_n\QM) \in \C'}

@@

%%T-Intf
 \inferrule[(T-Intf)]{
IT(\C) = \ann\ \QM{interface}\ \C\ \QM{extends}\ \C_1\ldots\C_n \
\oC\methods\ \cC\\\\
 \forall \QM{default}\ \mh\ \oC\QM{return}\ \e\QM;\cC \in \methods,
\ \ \Gamma^{\mh},\,\QM{this}{:}\C\vdash\e\in \_\subtype\C^{\mh} \\\\
 \forall \QM{static}\ \mh\ \oC\QM{return}\ \e\QM;\cC \in \methods,
\ \ \Gamma^{\mh}\vdash\e\in \_\subtype\C^{\mh} \\\\
\dom(\C)=\dom(\C_1)\cup\ldots\cup\dom(\C_n)\cup\dom(\methods)
 }{
\C \text{ OK}
}
\end{array}$
\caption{CJ Typing}
\label{ET}
\end{figure}

In Figure~\ref{ET} we show the typing rules.  We discuss first the
most interesting rules, that is \rn{t-Obj} and \rn{t-Intf}. Rule
\rn{t-Obj} is the most complex typing rule. Firstly, we need to
ensure that all field initializations are type correct, by looking up the type of
each variable assigned to a field in the typing environment and verifying that such type is a
subtype of the field type. Secondly, we check that all method bodies are
well-typed. To do this the enviroment used to check the method body
needs to be extended appropriately: we add all fields and their types;
add $\this:I$; and add the arguments (and types) of the respective
method.
Now we need to check if the object is a valid extension for that specific interface.
This can be logically diveded in two steps:
First  we check that all method headers are valid
with respect to the corresponding method already present in $\C$;

\noindent$\begin{array}{l}
\InTextDef{10em}{\sigvalid(\mh_1\ldots\mh_n,I)
}{
\forall i\in 1..n\ \ \mh_i\QM;\subtype\mBody(\m^{\mh_i},\C)
}
\end{array}$

Here we require that for all newly declared methods, there is a
method with the same name defined in the interface $\C$, and that such method is a subtype of the newly introduced one. We formally define here subtyping between method, in a more general form that will be useful later.

\noindent$\begin{array}{l}
\InTextDef{18em}{
\T\ \m\oR\T_1\x_1\ldots \T_n\x_n\cR \subtype \T' \m\oR\T_1\x_1'\ldots\T_n\x_n'\cR
}{
\T\subtype \T'
}\\
\InTextDef{18em}{\method \subtype
\QM{default}\ \mh\,\mbox{\Q@\{return \_;\}@}
}{
\method\subtype\mh\QM;
}\\
\InTextDef{18em}{\QM{default}\ \mh\,\mbox{\Q@\{return \_;\}@}\subtype\method
}{
\mh\QM;\subtype\method
}\\
\end{array}$

Two method headers are subtypes if all the parameter types are the same and the return types are subtypes.
That is, we allows return type specialization as introduced in Java\marco{insert the version number}.
Default methods are subtypes if their method headers are subtypes.

Finally, all abstract methods in the interface (that is methods that need
to be explicitly overridden) have been
implemented. That is, we define a method with the same name.

\noindent$\begin{array}{l}
\InTextDef{11em}{\alldefined(\mh_1\ldots\mh_n,I)}{
\forall\m\mbox{ such that } \mBody(\m,\C)=\mh\QM; \exists i\in 1..n\ \m^{\mh_i}=\m}
\end{array}$

@@

The body of a method $\m$ contains all the relevant information with respect to that method, like the type of $\m$ as well as the modifier.
We use internally a special modifier $\conflicted$ to denote the case of two methods with conflicting implementation.\\*
\noindent$\begin{array}{l}
\InTextDef{15ex}{
\mBody(\m,\C_0)
}{
\override(\methods(\m),
\shadow(\m,\Aux{tops}(\Cs)))
}\\
\InTextWith{\metaVar{IT}(\C_0) =
\ \ann\ \QM{interface}\ \C_0\ \QM{extends}\ \Cs\ \oC\methods\cC}
\end{array}$

@@

\tops{} recover from the interface table only the ``needed'' methods, that is,
the non static ones that are not transitively reachable by following another, less specific, superinterface chain. Formally:\\*

\noindent$\begin{array}{l}
\InTextAssert{25ex}{
\method\in\tops(\m,\Cs)}\\
\InTextWith{\C\in\Cs\ ,\method=\mBody(\m,\C), \method\mbox{ not a static method and } \nexists \C'\in\Cs\setminus\C \mbox{ such that } \C' \subtype \C}
\end{array}$

\shadow{} choose the most specific version of a method, that is the unique version available, or a conflicted version from a set of possibilities.
We do not model overloading, so it is an error if multiple versions are available with different parameter types. Formally:\\*
\noindent
$\begin{array}{l}
\InTextDef{15ex}{
\shadow()}{\none}\\
\InTextDef{15ex}{\shadow(\method)}{\method}\\
\InTextDef{15ex}{\shadow(\overline{\mh\QM;})}{\Aux{mostSpecific}(\overline{\mh\QM;})}\\
\InTextDef{15ex}{\shadow(\methods)}{\conflicted\mh\QM;}\\
\InTextWith{\methods\mbox{ not of the form }\overline{\mh\QM;}\mbox{ and }
\Aux{mostSpecific}(\methods)\in\{\mh\QM;,\QM{default}\ \mh\mbox{\Q@\{return \_;\}@}\}}\\
\end{array}$

Where $\Aux{mostSpecific}$ return the most specific method , that is a method whose type is the subtype of all the others.
Since method subtyping is a partial ordering, this may be not defined, this in turn makes \shadow{}, and the whole \mBody{} not defined for that specific $\m$. Rule  \rn{(t-intf)}  rely on this behaviour.

\noindent$\begin{array}{l}
\InTextDef{18ex}{\Aux{mostSpecific}(\methods)}{\method}\\
\InTextWith{\method \in \methods\ \mand\ \forall \method' \in \methods :  \method \subtype \method'}

\end{array}$

${}_{}$\\*
The override function models how the implementation in an interface can override implementation in the superinterface; even in case of a conflict.
Note how we use the special value $\none$, and how (forth case) overriding can solve a conflict.

\noindent$\begin{array}{l}
\InTextDef{18ex}{\override(\none,\none)}{\none}\\
\InTextDef{18ex}{\override(\method,\none)}{\method}\\

\InTextDef{18ex}{\override(\none,\method)}{\method}\\
\InTextWith{\method\mbox{ not of the form } \conflicted\ \mh\QM;}\\
\InTextDef{18ex}{\override(\method,\method')}{\method}\\
\InTextWith{\method'\in\{\mh\QM;,\QM{default}\ \mh\,\mbox{\Q@\{return \_;\}@}, \conflicted\ \mh\QM; \},
\method\subtype\method'}
\end{array}$

@@

@@

%\lstinputlisting[linerange=4-9]{../UseMixinLombok/src/methodshadowing/test5/Test5.java} % APPLY:linerange=JAVABUG
